---
title: "R Data Manipulation Tutorial: Carbon Capture Indicator"
author: "Tristan Misko"
date: "7 October 2022"
output: pdf_document
---
# Introduction

I've gotten a couple of questions about how to get started learning some of the
more high-powered tools like R and python and how to know when to use these tools 
over something simple like Google Sheets.  In most cases, I would recommend 
using Google Sheets when you can, but sometimes (as in the example below) it 
would be incredibly tedious to try to do the analysis by hand.  

In this document, I work through an example for which producing the indicator by
hand or in Google Sheets would be very difficult.  I collect and clean the
data for the Carbon Capture Indicator in the Land Category of the Sustainability Pillar.  
This example is great to learn from because it covers most of the issues you might
run into and uses a lot of the main tools you'd need to know to be able to 
manipulate and clean data to produce a useable indicator in R.  I've written this
tutorial to be accessible to anyone on the SSPI team, but if anything is unclear,
please reach out to me so I can fix it.

I'll probably also upload a regressions tutorial when I rerun the outcome variable
regressions in a few days, so look out for that.

# Crash Course on R Fundamentals

Computers store objects in memory and run programs which can manipulate those
objects into new objects.  Our goal is to learn some useful commands and workflows
for effectively manipulating dataframe objects.  The language we'll be using to
communicate which manipulations we'd like the computer to execute is called R.^[We'll 
be using R for a few reasons.  First, I and a number of your teammates have
some familiarity in R for data analysis, so there will be people around who can
support on your way up the learning curve.  Secondly, R is a free and open source
software used by analysts in academia and industry, so it's a valuable skill to 
develop.  Third, I think R is more beginner friendly than python for data 
analysis.]

R is a programming language designed for data manipulation and analysis, and it
is typically run in RStudio, an Integrated Development Environment (IDE) designed
for data manipulation.  RStudio will provide us a nice location to edit and run
our code and to see the outputs of our manipulations in real time.  To start up, 
you'll download R and RStudio following the instructions at the end
of the document.^[Follow this installation guide for more help: 
https://moderndive.netlify.app/1-getting-started.html]

R files (files ending in `.R`) store lines of R code.  Running such a file executes
the code, performing whatever actions author of the file has told R to do.  

R Markdown files (files ending in `.rmd`) are super nifty special files which
contain chunks of code you can run like an R file along with writing used to 
communicate what's going on in
the code or the larger analysis.^[`.r` is to `.py` as `.rmd` is to `.ipynb`] 
When you "knit" an `.rmd` file, the code
in the chunks runs, producing whatever data manipulations, tables, and graphs
you've instructed R to do, then the written document is compiled and returned.
You're actually reading the output of an R Markdown file right now. For the 
SSPI, `.rmd` files are great for communicating the idea of the kinds of more
complex analysis we'd be using R for in a way that's readable for people who
aren't used to writing code.

R contains objects and functions.  Typical objects for us will be values, vectors,
and dataframes:

* Values are numbers, stings of characters, etc.
* Vectors are ordered collections of values of the same type (i.e. all strings, all numbers)
* Dataframes are table-like collections of data values organized into columns 
(each column is a vector)

We will call R functions on these different types of objects to achieve the 
results we're looking for.  Functions take in **arguments** (objects to act on)
and **return** objects. 

Below, we'll see many examples of the types of objects
described above and we'll get to know many of the standard functions used to 
manipulate these objects.

One last note: R has a weird notation for variable assignment.  Most languages
would say `x = 5` to make a variable `x` which has value `5`.  You can do this
in R, but convention is to use the `<-` operator in R to assign variables.  That
is, instead of saying `x = 5` in R, you'd say `x <- 5`.  Weird, right?

# Setting Up the Document

At the top of an R Markdown file, we set up the file we're going to use.  We 
call the special `library` function to import standard libraries used in R.  A
library is a collection of code and documentations written by someone for other 
users.  This code defines some useful objects and functions we will use below.
Calling library will set up this outside code so we can use it as a base atop
which we will do our own work.

The last step is to set up our working directory.  This is the folder on our
computer in which we have our data stored.  Every time you import data in R,
you'll have to remember to set your working directory.  If ever you don't know
what your working directory is, you can call `getwd()` to figure it out.  To
set your own working directory, find the filepath to the folder with your data
on your computer, and call the `setwd` function on a string containing the 
filepath.^[Filepaths look different on MacOS and Windows, so watch out for that!]

```{r setup, results=F}
knitr::opts_chunk$set(echo = TRUE) # this line tells RMD to display the code
library(dplyr) # standardlibrary for manipulating data in R
library(ggplot2) # standard library for plotting data in R 
library(tidyr) # standard library we'll use to "spread" the data (see below)
library(rmarkdown) # a library for making pretty tables for the document
# set the working directory (the folder on my computer where i'm keeping the data and stuff)
setwd("/Users/tristanmisko/Documents/Berkeley/ECON/SSPI\ Research/Tutorials/2022-10-07\ R\ Data\ Manipulation\ Tutorial")
```

# Importing the Data

To start, we import the list of SSPI countries so that we don't have to type the 
name of each country by hand.  I created a nifty little `.csv` file for this 
from the main data file online.  You can easily create `.csv` file using 
`File > Download > Comma Separated Values (.csv)` on Google Sheets.

To import data, we call the `read.csv` function.  The string `"SSPI_countries.csv"`
is the filename of the data we want to import.  The `read.csv` function takes
the argument `"SSPI_countries.csv"` and returns a dataframe object containing
the data from that file.  We "pass" the dataframe returned by `read.csv` to 
the variable `SSPI_countries` using the assignment operator `<-`.  Now, whenever
we type `SSPI_countries` below, R knows we're talking about the dataframe we 
pulled down.  

```{r}
#"<-" is R's funny notation for "=".  I think "=" also works but I'm an R purist
SSPI_countries <- read.csv("SSPI_countries.csv")
# look at the first few rows of SSPI_countries dataframe object 
head(SSPI_countries)
```

Notice that the `SSPI_countries` dataframe has a lot of whitespace in it because
I was sloppy in making the `SSPI_countries.csv` file.  To get rid of that, we're 
going to `select` the second and third column of the dataframe.  We use the 
weird R notation `c(2,3)` to create (hence the `c`) the vector containing 2 and
3.  For our purposes, a vector is just a list of values.  Each column and row of
a dataframe is a vector, and there are special ways to access each of those in R.
More on that later.

The function `select` takes two arguments: our dataframe `SSPI_countries` and
the vector containing 2 and 3.  Calling `select` on the vector containing 2 and 3 
tells R to select the data that is in the second and third column of our dataframe, 
and to return a dataframe containing only that data in a new dataframe.^[R 
uses 1-indexing for various reasons. That is, list indices
are $1,2,3,\ldots$.  Most programming languages use 0-indexing (lists go 
$0,1,2,\ldots$) for various other reasons, and programmers get very upset about 
this kind of stuff.  For us, it's not a big deal, but if you're used to python
or something like that, just remember R is weird.] We then assign that returned
value to the variable `SSPI_countries`. 

If that seems a little weird at first, that's because it is. Variable names behave
differently on the left and right sides of the assignment operator `<-`.  Remember:
*everything to the right is evaluated (i.e. all the calculations and function 
calls and returns) before anything on the left side.*  In the first line of 
code below, evaluating the right side means that we access the original dataframe 
`SSPI_countries` in memory (i.e. the one with all the whitespace), we do our 
selection (which returns a new dataframe without any of the whitespace), then
we tell R to make `SSPI_countries` "point at" the clean new dataframe instead
of the old one.  Below this line, the variable `SSPI_countries` has in a sense
"forgotten about" the old dataframe with all the whitespace---it's no longer in
the computer's memory---and we only remember the clean new dataframe.  We make
use of this kind of operation (called "overwriting a variable") constantly in 
this file, so it's worth taking a second to make sure that's clear before moving on.  

```{r}
# use the select function from the dplyr library to eliminate the whitespace
SSPI_countries <- select(SSPI_countries, c(2,3))
#see how we removed the whitespace
head(SSPI_countries)
```

Now we're going to pull in the data we need for the indicator. We use the `nrow`
and `ncol` functions to see that the dataframe has 7329 rows and 14 columns.  We
view the names of the columns with `colnames` and decide that we really only need
the columns `"Area"`, `"Year"`, and `"Value"`. We then 
use our friend the `select` function to pick only these columns.  One nice thing
about `select` is that it can use indices (like `c(2,3)` above) or column names
from our dataframe (like below), and R will figure out which we mean.  
We save our selection over the original `carbon` dataframe and look at the result.

```{r}
# import the data
carbon <- read.csv("FAOSTAT_data_en_9-28-2022.csv")
# check the dimensions
c(nrow(carbon), ncol(carbon))
# check column names to select the ones we're going to need
colnames(carbon)
#select the right columns
carbon <- select(carbon, c("Area", "Year", "Value"))
head(carbon)
```

We'll also pull in the forest extent data using the same method. We can overwrite
the `colnames` of the dataframe with whatever we want by reassigning `colnames(forest)`
to a vector of the appropriate length containing our desired names. We can use 
`select` with a negative condition too: below we `select` the columns which are 
not (`!`) `FL_2010`, `FL_2016`, `FL_2017`, and `FL_2019`. 

```{r}
# import the forest data
forest <- read.csv("fra2020-extentOfForest.csv")
#look at the column names
colnames(forest)
#rename them looking at the online database so we can tell what we're talking aobut
colnames(forest) <- c("Area", "FL_1990", "FL_2000", "FL_2010", "FL_2015", "FL_2016", "FL_2017", "FL_2018", "FL_2019", "FL_2020")
# select only the potentially relevant years
forest <- forest %>% select(!c("FL_2010", "FL_2016", "FL_2017", "FL_2019"))
head(forest)
```

# Carbon Stock Data
## Cleaning and Filtering the Data

Next up, we're going to filter down the `carbon` data to only the SSPI countries using
the `SSPI_countries` dataframe we imported earlier.  To achieve this, we use the
aptly named function `filter` from the `dplyr` library.  The `filter` function
requires a logical vector to work, i.e. for each row in the data it needs to
know whether that row is in (`TRUE`) or out (`FALSE`).

For our purposes, we want to know whether the `Area` value in the dataframe (which
indicates the country name) is in our list of SSPI Countries.  We use the 
special vector operator `%in%`, which checks the element on the left of `%in%` is
contained in the vector on the right of `%in%`. We make the vector `countries` 
using the special `$` notation for dataframes.  The notation `SSPI_countries$Country`
tells R to look into the `SSPI_countries` dataframe, grab the column `Country`, 
and stick the result into a vector, which we've chosen to call `countries`.  
This vector has `nrow(SSPI_countries)`=`r nrow(SSPI_countries)` entries, one for 
each row of the dataframe.  

There are a couple of subtleties to watch out for when doing this with country
names.  Usually we prefer to use the three letter country codes standardized by 
the UN when we can because **some countries have multiple names**. In this data, 
there are no country codes, so we have to make do with names. The usual multiple name
suspects in the SSPI are 
Korea, Rep. (aka. South Korea, Republic of Korea, or simply Korea), Russian
Federation (often just listed as Russia), the United States (which sometimes
goes by United States of America, US, and USA depending on the dataset), and 
Slovak Republic (aka. Slovakia). It's always important to check that you've got
all the countries once you've done a big filtering operation.  To do this, we 
tell R to list out all of the unique `Area` names after filtering down to the
countries whose names are contained in the vector `countries`. 

```{r}
# make a vector containing the SSPI_country Country names
countries <- SSPI_countries$Country
# list out the unique country names in the filtered datset
unique(filter(carbon, Area %in% countries)$Area)
```

We can see that we're missing seven countries.  As we find below, we have the 
usual suspects and a few others have alternate spellings.  For these, we look 
through the data and see how they are encoding the names, then we simply add those 
names to our `countries` vector so that when we run the filter again, `%in%` will
return `TRUE` instead of `FALSE`.

```{r, results=F}
# check what the right names for the SSPI countries are
unique(carbon$Area)
# add the name used in the data to the list
countries <- c(countries, "Republic of Korea", 
               "United Kingdom of Great Britain and Northern Ireland",
               "United States of America", "Slovakia", "Türkiye", "Czechia")
# filter again and check to see that we have 48/49 countries
unique(filter(carbon, Area %in% countries)$Area)
```

Now only Kuwait is missing.  This makes sense: there aren't any forests in 
Kuwait! We'd deal with this issue in a meeting with the Professor.

Now we can run our final filter on the data, overwriting the original `carbon`
dataframe with only the data we need.  We have now reduced our original 
$7329\times 14$ dataframe full of irrelevant countries and variables to a much
more managable $732\times 3$ dataframe containing only the countries and variables
we actually want. This process gets referred to as "cleaning" or "selecting" and
"filtering" the data.  

```{r}
# run the filter for countries
carbon <- filter(carbon, Area %in% countries)
# run the filter for years
carbon <- filter(carbon, Year <= 1999 | Year >= 2015)
# check out how many observations we have left
nrow(carbon)
```

## Grouping and Averaging the Data

Now we're ready to compute 1990s averages.  General strategy: define a new variable
called `modYear` which equals "1990s" if the year is in the 1990s, otherwise it's 
just the year.  We can then call `group_by` and `summarize`, two special and 
especially confusing `dplyr` functions, which we'll use to group the observations
into buckets based on their `modYear` and `Area`.  For years outside of the 1990s,
there will only be one observation in the bucket: we only have observation for each
country in each year, so the average of that one observation will just be the 
value for the given year. For years in the 1990s, there are ten observations for
each country in the 1990s, all with a value of `"1990s"` for modYear, so computing
the average across a group for a given `Area` and `modYear == "1990s"` will 
return the average of all ten observations in the group.

This step is definitely not trivial, and there are many potential approaches.  
This one might not be the clearest, but it's important to get some exposure to
`group_by` and `summarize` because they are some of the most useful functions
for data manipulation in R (well, actually in `dplyr`).

```{r}
# define the modYear function which takes in a vector of years and returns the vector
# of years with all 1990s values replaced by "1990s"
modYear <- Vectorize(function(year){
    if (year <= 1999 & year >= 1990){
        return("1990s")
    } else {
        return(as.character(year))
    }
}, "year")

# apply the modYear function to the Year column of carbon, and store the results
# in a new column of carbon called modYear
carbon$modYear <- modYear(carbon$Year)
#look at the the table of carbon to check that we got the expected result
head(carbon)
# use the group_by and summarize operators to make a new dataframe which does 
# the averaging and removes the irrelevant years
carbon <- carbon %>% group_by(Area, modYear) %>% summarize(Value = mean(Value, rm.na = T))
head(carbon)
```

Above we used the `dplyr` pipe operators `%>%`, which might look scary at first
but which are actually quite nice.  The way to read the last line to the right of
the `<-` assignment operator is "take the `carbon` dataframe, send it into the 
`group_by` function, grouping observations which have the same `Area` and `modYear` 
value, then send that grouped data into the `summarize` function, which makes a 
new grouped dataframe which has the group identifier variables `Area` and `modYear`
and assigns `Value` (in the new grouped dataframe) to be the `mean` of `Value` across
the group (where `Value` is taken from the original `carbon` dataframe). 

Looking at `head(carbon)`, we can now see that we have a 1990s average and all
the years following 2015. Sweet! 

## Spreading to Do Computations

We now want to compute percent changes, which is going
to require us to reorganize the dataframe.  In particular, we want to have a single 
row for each `modYear` so that we can easily compute the averages for all countries
at once using column operations.  

To achieve this, we're going to use the `spread` function from the `tidyr` 
library.   ^[When people say R is an "ugly" language, this is what they 
mean.  Basic things like spreading data require the use of an outside libary, 
and the names of functions across the libraries are not consistent and sometimes
conflict.  A "pretty" language would have all this built-in and standard, but that
can add complexity.  It's all tradeoffs :/] Specifically, we are going to 
spread `Value` across a different column for each value of `modYear`.  

In the language that the `spread` function uses, `Value` is the "value" and `modYear`
is the "key".  The idea of using "key-value pairs" to identify objects and 
define functions is one of the most important in computer science.  Within each
`Area`, we have a set of keys (each observation of `modYear`) which each correspond
to a single `Value`, and we're going to spread the keys across the columns by
calling `spread(dataframe, key, value)`. 

```{r}
# run the spread function and store the value as carbon
carbon <- spread(carbon, modYear, Value)
head(carbon)
```

A quick look at the data shows that we've almost got everything we want, but
there's an issue with Belgium and Luxembourg: they don't have a 1990s average!
What's going on? To find out, we've got to go back and look at our original data, but
that's all been overwritten. ^[For you reading this document, you can
actually still see the original data in the paged output at the top of the sheet.  
Can you spot the issue there before reading on?]

What might we do? Well, we can import the original data again into a separate dataframe 
and look at Belgium and Luxembourg values. Paging through the data is one solution,
but I'm going to use some `regex` to find the issue. 

We use `regex` whenever we are trying to match strings.  The term `regex`, a
portmanteau for "regular expressions," refers to a collection of procedures 
that dates back to the early days of computer science.^[A bit of `regex` history https://www.youtube.com/watch?v=NTfOnGZUZDk] Fun fact: I've actually used `regex` 
in every research project that I've done, so it's definitely a useful little 
topic to master.  

For our purposes, we'll use the function `grepl(pattern, string)`, which checks
whether or not `string` contains `pattern`.  There are a lot of special ways to
make patterns to account for very complex cases, but for us we'll just use the
country names we're trying to find.  In R, `grepl` is vectorized, which means
that we can seach across a vector of strings (like the column `fix_bnl$Area`) and
the function will return a vector of the same length containing `TRUE`s and `FALSES`
in the appropriate positions.^[When defining the `modYear` function above, I 
manually vectorized using `Vectorize` it so it could do a whole column at once.
Any user-defined function can be vectorized using 
`Vectorize(userdefined-function, "vectorized_variable")`]

```{r}
# import the data into a new dataframe
fix_bnl <- read.csv("FAOSTAT_data_en_9-28-2022.csv")
# define a logical vector that is true if the Area name contains Belgium or Luxembourg and
# stor it as a column in fix_bnl dataframe
fix_bnl$contains_bnl <- grepl("Belgium", fix_bnl$Area) | grepl("Luxembourg", fix_bnl$Area)
# select the relevant columns and then filter for the columns which contain_bnl
fix_bnl <- fix_bnl %>% select(c("Area", "Year", "Value", "contains_bnl")) %>%
    filter(contains_bnl)
# look at the data
head(fix_bnl)
```

We can see that Belgium and Luxembourg get lumped together for the 1990s.  This
now presents us with a few options, which we would likely have a meeting with 
the Professor and potentially an expert.

1. We could throw out the countries and say we don't have data, but we could 
almost certainly do better.
2. We could go looking for better data, but we're probably not going to find any.
This is a UN dataset with tons of information for most countries.  If the UN
couldn't find the data, you probably can't either.
3. We could just give Belgium and Luxembourg the same combined values as their
1990s average, but this would penalize both countries because they're comparing
to too high a baseline.
4. We could try to **impute** a value making a few reasonable assumptions about
the data.  This is the approach we take below.

## Imputing Values for Belgium and Luxembourg

So, we've got combined Belgium-Luxembourg data through 1990-1999.  To make a 
1990s average for Belgium and Luxembourg, the approach that seems best to me
is to make the average 1990s average for the combined data, then to divide 
that average between Belgium and Luxembourg based on the 2000 proportions, which
is the first year for which we have data.  This might not be exactly right, but
if carbon stock in forest area is proportional to land area, then it should be
pretty close to right.  Mostly, it seems to be about as good as we can do
with the data we have, so we'll procede with caution.

To determine whether this is even a good idea, we might look to see whether the 
ratio between Belgium and Luxembourg is roughly constant over time, which would justify
this kind of step.  To do this, we're going to group the observations by `Year`
(using `group_by`) and `summarize` the quotient of the `first` `Value` in the group
(the `Value` corresponding to Belgium in the given group `Year`) and the `last`
`Value` in the group (corresponding to Luxembourg).

```{r}
ratios <- fix_bnl %>% group_by(Year) %>% summarize(ratio = first(Value)/last(Value))
head(ratios)
```
Looking at the output
of this grouping, which we've stored in the grouped dataframe `ratios`, we can
see that the ratio is declining over time, but not by that much.  This means that
Luxembourg's carbon stock in forests has been increasing faster than Belgium's
over the last 20 years.  

With this information, we have to make a decision.  Do we want to use the average
ratio over time or the ratio in 2000? 

* Using the average ratio would imply that we think
that the ratio in the 1990s could have been higher or lower than it was in 2000,
and we don't really have any better information. 
* Using the 2000 ratio would imply
that we think there's a pattern over time and that using 2000 carries more information
about what the likely ratio was through the 1990s than the average through the
2000s and 2010s. 

I personally think the second option is better because of the aforementioned 
steady decline in the ratio over time, but because they're both between 8 and 9
it won't really matter that much. We could quantify that intuition using a 
sensitivity test if we wanted, but for now I'm going to run with the 2000 ratio.

To find the appropriate weights, we're going to go back to the `fix_bnl` dataframe 
and find the 2000 value for Belgium and Luxembourg using `filter`.  Assigning the 
result to `bl2000`, we can now pull out the `Value` column into a vector and divide 
this vector by the `sum` across the vector to obtain a new vector containing the 
proportions of the 2000 total across Belgium and Luxembourg held by Belgium and
Luxembourg respectively.

```{r}
# filter the fix_bnl dataframe to be only year 2000
bl2000 <- fix_bnl %>% filter(Year == 2000)
# check to make sure we filtered correctly
head(bl2000)
# compute the weights: noitice when we divide and multiply vectors by constants,
# all the elements of the vector get divided/multiplied by that constant.
w <- bl2000$Value/sum(bl2000$Value)
# look at the weights: Belgium is much bigger, so it gets higher weights (checks out)
w
```

Next, we compute the 1990s average.  By now you probably know the drill: `filter`
down the data, pull out the column, record the mean.
```{r}
# filter data down to combined Belgium-Luxembourg values ranging 1990 to 1999 only
bnl_1990s <- fix_bnl %>% filter(Area == "Belgium-Luxembourg")
# compute the mean value across the 1990s for combined Belgium-Luxembourg
bnl_1990s_mean <- mean(bnl_1990s$Value)
# print out the Belgium-Luxembourg 1990s mean
bnl_1990s_mean
```

Finally we find the imputed values by multiplying the 1990s mean by the weights:
```{r}
# multiply the weight vector w by the 1990s mean
imputations <- bnl_1990s_mean*w
# print out the result
imputations
```
As a final sanity check, we look back at the 2000 values to see that our imputations
aren't so far off.  Sure enough, the 2000s values are a bit larger for Belgium
and Luxembourg than the 1990s average, but we also know that the combined Belgium-Luxembourg 
average was growing through the 1990s, so these numbers seem pretty good.

We now want to stick these values into our `carbon` dataframe in the appropriate
slots.  To do this, we'll use a concept called *logical indexing* or 
*logical subsetting*. One nice feature of R vectors is that you can select items 
from a vector $v$ using a logical vector of the same length as $v$.  This is 
*incredibly* useful when working with big datasets, especially for filling in 
missing values.  

What we want to do is assign our imputed values to the correct positions in the
`carbon` dataframe, i.e. the positions which currently have an `r NA` value. To
find the elements which are currently `NA`, we use `is.na` called on our 
vector (which, recall, is the 1990s column of the `carbon` dataframe).  This
returns a logical vector of the same length as `carbon`'s columns which we can
use to do subsetting.  

As in most programming languages, subsetting in R is
carried out using `[]`.  For example, if `a <- c(2,3,4)`, then `a[1]` is `2`
and `a[c(TRUE, FALSE, TRUE)]` is the vector `c(2,4)`.^[Really it's the vector 
containing 2 and 4.  The command `c(2,4)` is what the user would use to create 
such a vector, but it isn't the name of/notation for the vector itself.  Unfortunately,
there's not really a standard written notation for an R vector, so
we're stuck with boring footnotes and bad notation.  Just another way in which
R is "ugly"]. Doing logical subsetting for our NAs then putting in an assignment
statement will overwrite the elements selected in the logical subsetting with whatever you assign them to be on the right hand side.

```{r}
#our logical vector: backticks are used to specify numeric column names
is.na(carbon$`1990s`)
# using our logical vector to select the appropriate elements of our vector
carbon$`1990s`[is.na(carbon$`1990s`)]
# using our logical vector to overwrite the appropriate elements of our vector
carbon$`1990s`[is.na(carbon$`1990s`)] <- imputations
# check to see that our fix worked (huzzah!)
head(carbon)
```
Now the hole in the data is fixed!

# Forest Data

We now `filter` the forest data to eliminate the non-SSPI countries.  I'll go 
much quicker in this section.  See if you can follow my code. 

```{r}
# create a logical vector to see which countries are missing
check_inclusion <- unique(countries) %in% unique(filter(forest, Area %in% countries)$Area)
# present the missing countries in a human readable format
head(arrange(data.frame(Area = unique(countries), check_inclusion), Area))
```

We're missing a three countries: UAE, Kuwait, and Saudia Arabia.  Once again, 
we see that the Middle Eastern countries which are mostly in desert climates
don't fit well with our forest-based indicator.  We make a new dataframe and
`rbind` (i.e. rowbind) it to the end of our real dataframe, then run the final
filtering, checking that we get all 49 countries.

```{r}
#make an additional dataframe, defining the column names and the vectors they contain
additional_data <- data.frame(Area = c("Kuwait", "Saudi Arabia", "United Arab Emirates"), 
                              FL_1990 = c(0,0,0),
                              FL_2000 = c(0,0,0),
                              FL_2015 = c(0,0,0),
                              FL_2018 = c(0,0,0),
                              FL_2020 = c(0,0,0))
# check out our new extra dataframe
head(additional_data)
# rowbind the additional zeros into the original dataframe
forest <- rbind(forest, additional_data)
# see that we've done what we expected
head(forest)
# run the filter and overwrite forest
forest <- forest %>% filter(Area %in% countries)
# check that the number of rows is right
nrow(forest)
```

# Merging `carbon` and `forest`

Ultimately, we want everything in a single dataframe so that we can do 
computations easily.  We will execute a `merge`, matching each observation (i.e.
each row) based on the `Area` column.  The strings must match exactly, otherwise
the `merge` won't work.  Luckily, the country names match up because both 
dataframes are from the same source, FAO.  

```{r}
# create the merged dataframe
carbon_capture <- merge(carbon, forest)
# check that it looks right
head(carbon_capture)
```

And just like that, we've got everything we need! We'll just do a quick cleanup,
renaming the columns with `colnames` so that we know what's what and dropping the 
2015, 2016, 2017, and 2019 columns from the `carbon` dataframe using `select`. 

```{r}
# drop unnecessary columns
carbon_capture <- carbon_capture %>% select(!c("2016","2017","2019"))
# rename the carbon columns so we know what they are
colnames(carbon_capture)[2:5] <- c("C1990s", "C2015", "C2018","C2020")
# view the resulting dataframe to make sure everything's looking good
head(carbon_capture)
```


# Finishing Up the Computations

Finally, we can do the computations for the indicator.  From the indicator table,
we know we are trying to get the "percentage change in the ratio of carbon stock
in living biomass over forest land from a 1990s average."  First, we compute an
approximate 1990s average for forest land by averaging the 1990 and 2000 value.
Next, we compute the ratios for 1990s and 2018.  Finally, we compute the percentage
change in the ratios.  All that work above for a couple lines of code down here.

```{r}
# compute an average of forest land for 1990s
carbon_capture$FL_1990s <- (carbon_capture$FL_1990 + carbon_capture$FL_2000)/2
# compute the ratio of carbon stock per forest land for the 1990s
carbon_capture$r1990s <- carbon_capture$C1990s/carbon_capture$FL_1990s
# compute the ratio of carbon stock per forest land for 2018
carbon_capture$r2018 <- carbon_capture$C2018/carbon_capture$FL_2018
# compute the percentage change in ratios
carbon_capture$pchg_1990s_to_2018 <- (carbon_capture$r2018 - carbon_capture$r1990s)/carbon_capture$r1990s * 100
```

# Exporting the Data



Once you've done your computations, use `write.csv` to create a new `.csv` file
in the working directory with your data in a sharable format. Huzzah! We have the
raw data for an indicator ready to go.

```{r}
# write the dataframe to a new csv file which we can upload
write.csv(carbon_capture, "carbon_capture.csv") 
```

# Getting Your Hands Dirty

## Getting Started with R

1) Download R and RStudio and follow the setup steps.
2) Immediately change the Editor Theme to "Chaos" in the Appearance tab of Preferences
so that you don't sear your retinae with the white background.  
3) In the console, type `"install.packages(c("dplyr", "ggplot2", "tidyr", "rmarkdown"))`,
which will pull down all of the packages we need for R from CRAN (an internet repository) and 
store them on your machine.  If you don't do this, then calling, say,
`library(dplyr)` will send your computer on a wild goose chase looking for a
package you don't have installed.
4) Download the `r-for-SSPI-tutorial-1` directory from the google drive, and store
it in a folder you know the filepath to.  You see my filepath is long because
it's stored under many layers of folder organization.  If you're having trouble,
try putting it on your Desktop, which usually has a fairly simple filepath. 
5) Set your working directory to the folder you've just downloaded by changing the
filepath in the `setwd` function in line 17 to whatever the right filepath is 
for your machine.  If you're working on Windows, maybe do some googling to see
what the correct formatting is for non-Unix (which is to say silly---don't @ me) 
filesystems.

## A Note on Why We're Doing This 

High powered tools like R and python are great because they are **reproducable** 
and **extensible**.  For a research team like ours, reproducability is key because 
we want to be able to see what we've done and catch any mistakes along the way.  
Writing code to do our cleaning/analysis (as opposed to doing it in Google Sheets or Excel)
means that we have a recorded set of instructions to reproduce the analysis and
we can easily check to see if we've made any mistakes along the way. The exercises
below will demonstrate why extensibility is great.

## Exercises

1. Compute the percentage increase from 1990s to 2020 and store it in a separate
column called `pchg_1990s_to_2020` using the same technique we used to compute
`pchg_1990s_to_2018` above.  

2. Repeat this analysis to include the SSPI Extended Countries.  How do we change which
countries are included in the analysis? What file needs to be changed?  Use this
file to get an output dataset which has the original SSPI 50 countries (actually
48 for us) and the 18 extended countries.  Troubleshoot any issues that crop up
with country names, missing data, etc.

3. Right now we are using a 1990s (1990-1999) average and comparing it to 2018 data.  
If we want to get a comparable indicator for 2020, we might want to consider using the
average from (1992-2001) as the baseline.  Change the code I've provided so that
the only `modYear`s we group by are `"1992-2001"` and `"2020"`, and return a
new version of the final dataframe with this new version.  

*Find time to work as a group on these exercises and help each other!  If you're
having trouble, try googling it.  Usually hundreds of other people have had the
same issue.*

*If you're really stuck, email me at tjmisko@berkeley.edu with your questions. 
Finally, remember, you can always download this file again if you feel like it's 
gotten messed up.*